Two subtle problems with over-representation analysis

### 2024-09-16 manuscript/manuscript.Rmd
ms work

To quantify the effect of these two problems on RNA-seq based enrichment analysis, data from seven
transcriptomic studies were downloaded from DEE2 using the getDEE2 R/Bioconductor package
[@Ziemann2019-ux].
The criteria for selection included human as the species, presence in the DEE2 database,
passing most DEE2 quality control measures, a simple control-case experiment design with
two or three replicates per group, and with 50 or more differentially expressed genes
(FDR<0.05).
Raw data for these seven studies are available from NCBI Sequence Read Archive under the
accession numbers SRP128998, SRP038101, SRP037718, SRP096177, SRP097759, SRP253951 and
SRP068733 [@Felisbino2021-hc; @Lund2014-ds; @Rafehi2014-rv; @Keating2014-ab;
@Blanco-Melo2020-hh; @Rafehi2017-qn].
For each dataset, kallisto (v0.43.1) [@Bray2016-lu] transcript counts were aggregated to the
gene level.
Genes with fewer than 10 reads per sample on average were removed from downstream analysis.
Remaining genes passing this selection were included in the background gene set.
Differential expression analysis was conducted with DESeq2 v1.44.0 [@Love2014-wg], with
genes identified by their Ensembl identifiers.
Gene symbols were then fetched using biomaRt v2.60.0, based on Ensembl version 112 [@Durinck2009-dr].

@@

RNA-seq counts for SRA run accession ERR2539161 were obtained from DEE2 [@Mo2018-zb].
This dataset was selected due to its high number of assigned reads (367M).
A library of 200 random gene sets, each containing 30 genes, was generated.
To generate differential expression profiles, 5% of gene sets were selected to be
differentially expressed, with equal numbers of up- and down-regulated pathways.
Pseudosamples were generated by first downsampling the counts using the `thincounts`
function of the `edgeR` package v4.2.0 [@Robinson2010-eb] to 20M reads, followed by the
addition of some extra variation drawn from a normal distribution with mean of 1 and
standard deviation varying between 0.1 and 0.6, this value selected randomly for each gene.
Three "control" and five "case" pseudosamples were generated.
For the genes selected to be differentially expressed, expression values were multiplied by a
log fold change in the case samples of 0.5 for upregulated genes and -0.5 for downregulated
genes.
Genes that were selected to be both up and downregulated were left unchanged.
With these expression profiles, DESeq2 was used for differential analysis, and the results
were subjected to clusterProfiler which suffers from both problems,
clusterProfiler with the background problem fix, clusterProfiler with the FDR problem fix,
clusterProfiler with the fixes to background and FDR problems, an ORA function called `fora`
from `fgsea` bioconductor package v1.30.0 that does not suffer these problems, and `fgsea` a
fast implementation of the preranked gene set enrichment analysis (GSEA) algorithm
[@Korotkevich2016-gd].
For the ORA functions above, significantly differentially expressed genes with FDR 0.05 were
selected for the foreground gene set.
If fewer than 200 genes were detected as significant, then 200 genes with the smallest
p-values were used for ORA, like a previous report [@Tarca2013-xh].
For `fgsea`, the DESeq2 test statistic was used for scoring differential expression.
After enrichment analysis with these approaches, the significant up and down regulated
sets were selected and compared to the ground truth to calculate precision, recall
and F1 score.
At each value of added variance, 1000 replications were conducted.
No set seed was used, but results were similar when run on different computer systems.
All analyses were undertaken with R v4.4.0 in a Docker container.

@@

A workaround to eliminate the background problem was developed and we used it to compare the
results of ORA of seven datasets with the original and corrected methods for nine different
gene set libraries.
The mean number of statistically significant results was calculated for each of the gene set
libraries before and after correcting the problem (**Figure 1A**).
Each of the gene set libraries showed an uplift in the number of significant gene sets when correcting the background problem.
The smallest increase was observed in Cellmarkers gene library (11.8%), while the largest
increase was observed for HPO (224%).
The similarity between results, as quantified with the Jaccard similarity, was highest for
Cellmarkers and GTRD at 0.76 and 0.77 respectively, and the lowest similarity was recorded for
HPO and KEGG with 0.20 and 0.48 respectively.
Observed Jaccard scores correlated with the number of genes annotated to one or more functional
categories in the gene set library (**Figure 1B**).
This result suggests that ORA with smaller gene set libraries like HPO and KEGG are impacted
more severely.

![**Figure 1A. Effect of correcting the background problem.** Figures represent the mean number
of statistically significant sets (FDR<0.05) over seven independent datasets detected with
clusterProfiler with and without correcting the background problem.
Separate tests for up- and down-regulated gene lists were conducted.
The percent increase in significant sets are shown.](../analysis/fig1_ORAsig.png "Figure 1A
shows that correcting the background bug improves the number of significant sets between 12 and
229%.")

![**Figure 1B. Impact of the background problem is worse for smaller gene set libraries.** Mean
Jaccard similarity index was calculated for original and corrected ORA for seven independent
datasets. The number of genes represented in one or more functional sets is shown on the x-axis,
and the mean Jaccard index is shown on the y-axis.](../analysis/fig1_ORAjac.png "Figure 1B
shows that the background bug impacts smaller gene sets libraries like KEGG and HPO more
severely.")

@@

A workaround for the FDR problem was developed, then original and corrected
analyses were conducted for the seven datasets and nine gene set libraries.
Results showed no major differences in the number of significant sets between
original and corrected analyses (**Figure 2A**).

@@

![**Figure 2A. Effect of correcting the FDR problem.** Figures represent the mean number of
significant sets with and without correction of the FDR problem over seven independent RNA-seq
datasets. Foreground list consisted of 2000 genes with the smallest p-values](../analysis/fig2_ORAfdr.png "Figure 2A shows that correcting the FDR bug slightly reduces the number of significant sets.")

![**Figure 2B. Impact of the FDR problem is worse for shorter foreground gene lists.** Mean Jaccard
index between original and corrected analysis were calculated for seven RNA-seq datasets and nine
gene set libraries. Only GO, Reactome, Cellmarkers and TFT GTRD are shown.](
../analysis/fig2_ORAsz.png "Figure 2B shows that the impact of the FDR bug is worse when
foreground gene lists are shorter than 500.")

@@

Simulated differential expression profiles with *a priori* changes to predetermined pathway
genes were analysed in parallel by clusterProfiler default and with mitigations for
problems described above.
In addition, these data underwent ORA with fora which does not suffer from these two
problems.
For comparison, fgsea was used to compare the performance of these two ORA methods against a
prototypical FCS method.
These methods were tested with different levels of introduced variation and the mean
precision, recall and F1 scores were recorded (**Figure 3A**).
Mitigating the background problem improved recall (by 1.8%), however it caused
a similar-sized decrease in precision (1.6% lower).
Addressing the FDR problem improved precision by 9.9%, but the recall dropped by 7.7%.
Addressing both problems improved precision by 8.2% but decreased recall by 5.4%.
Under these conditions, the FDR problem affected results more severely than the background
problem.
Results obtained for clusterProfiler after implementing the two mitigations were identical to
fora to three significant figures.
fgsea recorded precision of 0.940, similar to the ORA methods, but recall was far superior,
with fgsea scoring 22.9% higher than default clusterProfiler and 28.3% higher than fora.
As a result, fgsea's overall accuracy as defined by F1 index was 21.0% higher than fora and
18.6% higher than default clusterProfiler.

When looking at the accuracy profile at different levels of added noise (**Figure 3B**),
ORA techniques show high precision when the added noise is low (<0.2), but
precision drops at higher levels of added noise.
The rate of drop in precision was more severe for approaches that suffer from the FDR problem.
The precision of fgsea was relatively stable despite high levels of added noise, ranging from
0.95 to 0.92.
The recall of fgsea was superior to all ORA approaches across all levels of added noise.

![**Figure 3A. Impact of correcting these two problems on accuracy of simulated expression
data.**
Precision, recall and F1 scores are the mean of trials at a range of added noise.
CP; clusterProfiler, CP BG fix; clusterProfiler with a workaround to the background problem,
CP FDR fix; clusterProfiler with a workaround to the FDR problem,
CD BG and FDR fix; clusterProfiler with mitigations for both background and FDR problems,
fora; ORA function of the fgsea package,
fgsea; an FCS method similar to GSEA.
](../analysis/sim/fig3_bars.png "Figure 3A shows that fora is more
precise than clusterProfiler, and fgsea has higher recall than ORA methods.")

@@

To understand whether FCS has better sensitivity than ORA in a real setting, we conducted a
systematic downsampling of a gene expression dataset consisting of normal and tumour samples
of 37 lung cancer patients.
In the full set of patient samples, fora (ORA) yielded 375 significantly differentially
regulated Reactomes, while fgsea (FCS) gave 470 (FDR<0.05).
Fgsea yielded consistently more significant pathways than fora in downsampled data
(**Figure 4A**).
Fgsea also identified a larger proportion of pathways in downsampled data compared to fora.
For example at a sample size of 5, fgsea identified 74.5% of pathways from the full set as
compared to 41.5% for fora, confirming better recall for FCS over ORA.
It was noted that fgsea identified more pathways in downsampled data that were
inconsistent with the full dataset as compared to fora (**Figure 4B**).
However, inconsistent pathways were a smaller proportion of all findings for fgsea as
compared to fora (**Figure 4C**), indicating an overall lower false positive rate for FCS
compared to ORA methods.

![**Figure 4. Downsampling confirms higher sensitivity for FCS over ORA.**
(A) The number of significant pathways identified in downsampled data that are consistent
with findings in the full dataset, using FCS (fgsea) and ORA (fora) methods.
The blue and red horizontal lines indicate the number of significant pathways identified
in the full set of 37 patients using fgsea and fora respectively.
(B) The number of significant pathways identified in downsampled data that are not
consistent with the full dataset.
(C) The proportion of pathways in downsampled data that are inconsistent with the full
dataset.
All values are the median of 50 simulations.](../analysis/fig4.png "Figure 4A shows that fgsea is more sensitive than fora. Figure 4B shows fgsea has higher absule number of false positives, but Figure 4C shows that as a proportion of all findings fgsea has lower false positive rate than fora.")

@@

The background problem came to our attention when a doctoral student in our department
was puzzled by results obtained from clusterProfiler when using a custom set of pathways.
This exact issue was the subject of several posts to online bioinformatics forums dating to
~2020 where confused users were wondering why the number of background genes changed so much.
The results shown here indicate the background problem affects smaller gene set libraries
like KEGG and HPO more severely as compared to Cellmarkers and GO libraries which describe
a larger proportion of all genes.
ClusterProfiler maintainers recently provided an option to prevent the loss of unannotated
genes from the background `options(enrichment_force_universe = TRUE)`, however this feature
isn't yet described in the official documentation.

@@

So end-users can explore the effect of these two problems on their work, we have developed
an R/Shiny web application (see "Availability of data and materials" below).
Users can upload their foreground and background lists, select their preferred gene set
library and compare results from default and patched analyses.

ClusterProfiler is not the only tool to suffer from these problems, they are widespread
(**Table 2**). Still, there are some tools free of such problems.
For example, fora is a good option for R-based workflows, however the function does not calculate
the fold enrichment score,
so users will need to do it themselves [@Korotkevich2016-gd].
ShinyGO is an excellent alternative for web based analysis [@Ge2020-lu].
It doesn't suffer from the two issues described here, it is easy to use,
and has a high degree of reproducibility thanks to its stand-alone versioned docker
container option for local execution.
One problem that all of these tools listed in Table 2 have is that a background list is optional.
Sometimes this feature is hidden under "advanced options".
Given the strong biological and technical biases that are present in contemporary omics like single cell and spatial transcriptomics,
it should be a mandatory step for users to provide a custom background list.

@@

These issues together with previously mentioned pitfalls such as lack of background
correction and methodological misreporting [@Timmons2015-ki;@Wijesooriya2022-fw], suggest
that if the data can be scored/ranked, methods like FCS are strongly recommended over ORA.
ORA is better suited to other cases where genes are classified into binary groups.
For example when investigating the over-represented categories in gene expression modules
[@Langfelder2008-hu], or with variant-harbouring genes associated with a disease [@White2019-ps].

>>>>

To quantify the effect of these two problems on RNA-seq based enrichment analysis, data from seven
transcriptomic studies were downloaded from DEE2 using the getDEE2 R/Bioconductor package
[@Ziemann2019-ux].
The criteria for selection included human as the species, presence in the DEE2 database,
passing most DEE2 quality control measures, a simple control-case experiment design with
two or three replicates per group, and with 50 or more differentially expressed genes
(FDR<0.05).
Raw data for these seven studies are available from NCBI Sequence Read Archive under the
accession numbers SRP128998, SRP038101, SRP037718, SRP096177, SRP097759, SRP253951 and
SRP068733 [@Felisbino2021-hc; @Lund2014-ds; @Rafehi2014-rv; @Keating2014-ab;
@Blanco-Melo2020-hh; @Rafehi2017-qn].
For each dataset, kallisto (v0.43.1) [@Bray2016-lu] transcript counts were aggregated to the
gene level.
Genes with fewer than 10 reads per sample on average were removed from downstream analysis.
Remaining genes passing this selection were included in the background gene set.
Differential expression analysis was conducted with DESeq2 v1.44.0 [@Love2014-wg], with
genes identified by their Ensembl identifiers.
Gene symbols were then fetched using biomaRt v2.60.0, based on Ensembl version 112 [@Durinck2009-dr].
Descriptive information of the seven datasets is provided in Table 1.

@@

RNA-seq counts for SRA run accession ERR2539161 were obtained from DEE2 [@Mo2018-zb].
This dataset was selected due to its high number of assigned reads (367M).
A library of 200 random gene sets, each containing 30 genes, was generated.
To generate differential expression profiles, 5% of gene sets were selected to be
differentially expressed, with equal numbers of up- and down-regulated pathways.
Pseudosamples were generated by first downsampling the counts using the `thincounts`
function of the `edgeR` package v4.2.0 [@Robinson2010-eb] to 20M reads, followed by the
addition of some extra variation drawn from a normal distribution with mean of 1 and
standard deviation varying between 0.1 and 0.6, this value selected randomly for each gene.
Three "control" and five "case" pseudosamples were generated.
For the genes selected to be differentially expressed, expression values were multiplied by a
log fold change in the case samples of 0.5 for upregulated genes and -0.5 for downregulated
genes.
Genes that were selected to be both up and downregulated were left unchanged.
With these expression profiles, DESeq2 was used for differential analysis, and the results
were subjected to clusterProfiler which suffers from both problems,
clusterProfiler with the background problem fix, clusterProfiler with the FDR problem fix,
clusterProfiler with the fixes to background and FDR problems, an ORA function called `fora`
from `fgsea` bioconductor package v1.30.0 that does not suffer these problems, and `fgsea` a
fast implementation of the preranked gene set enrichment analysis (GSEA) algorithm
[@Korotkevich2016-gd].
For the ORA functions above, significantly differentially expressed genes with FDR 0.05 were
selected for the foreground gene set.
If fewer than 200 genes were detected as significant, then 200 genes with the smallest
p-values were used for ORA, like a previous report [@Tarca2013-xh].
For `fgsea`, the DESeq2 test statistic was used for scoring differential expression.
After enrichment analysis with these approaches, the significant up and down regulated
sets were selected and compared to the ground truth to calculate precision, recall
and F1 score.
At each value of added variance, 1000 replications were conducted.
No set seed was used, but results were similar when run on different computer systems.
All analyses were undertaken with R v4.4.1 in a Docker container.

@@

A workaround to eliminate the background problem was developed and we used it to compare the
results of ORA of seven datasets with the original and corrected methods for nine different
gene set libraries.
The number of statistically significant results was calculated for each of the gene set
libraries before and after correcting the problem (**Figure 1A**).
Each study was affected to a similar degree (mean uplift 125% ±199%), except for #5 which saw the number of
significant GO terms increase from 8 to 54 after correction (**Table S1**).
The similarity between results, as quantified with the Jaccard similarity, was highest for
Cellmarkers and GTRD at 0.76 and 0.77 respectively, and the lowest similarity was recorded for
HPO and KEGG with 0.20 and 0.48 respectively.
Observed Jaccard scores correlated with the number of genes annotated to one or more functional
categories in the gene set library (**Figure 1B**).
This result suggests that ORA with smaller gene set libraries like HPO and KEGG are impacted
more severely.
We investigated discrepant results for data set #1 (SRP128998) downregulated genes in HPO gene sets.
A scatterplot of significance scores (**Figure 1C**) shows severe deviation in p-values in gene sets in the largest
quintile (109-1382 members detected), suggesting larger gene sets are more severely affected by the background
problem.

![**Figure 1A. Number of statistically significant pathways (FDR<0.05) in seven independent RNA-seq studies with and
without correction of the background problem.
Separate tests were conducted for up- and down-regulated gene lists.
Mean values are depicted as an “x”. The percent increase in the mean number of significant sets are shown.
“*” indicates that 1 has been added to all values before log transformation.](
../analysis/fig1_ORAsig.png "Figure 1A
shows that correcting the background bug improves the number of significant sets between 12 and
229%.")

![**Figure 1B. Impact of the background problem is worse for smaller gene set libraries.**
Mean Jaccard similarity index was calculated for original and corrected ORA for seven independent datasets.
The number of genes represented in one or more functional sets is shown on the x-axis, and the mean Jaccard index is
shown on the y-axis; whiskers represent the standard deviation.](
../analysis/fig1_ORAjac.png "Figure 1B shows that the background bug impacts smaller gene sets libraries like KEGG
and HPO more severely.")

![**Figure 1C. Significance values (-log10 p-values) for dataset #1 (SRP128998) enrichment analysis
using HPO gene sets with and without correction of the background problem.**
Gene sets were divided into quintiles based on the number of genes detected in the dataset.](
../analysis/fig1_HPOdn.png "Figure 1C shows that larger gene sets are affect the p-values more severely by the
background bug.")

@@

A workaround for the FDR problem was developed, then original and corrected
analyses were conducted for the seven datasets and nine gene set libraries.
Results showed no major differences in the number of significant sets between
original and corrected analyses (**Figure 2A**).
The impact of the FDR problem was similar across these seven datasets (**Table S2**).

@@

![**Figure 2A. Effect of correcting the FDR problem.**
Figures represent the mean number of significant sets with and without correction of the FDR problem over seven
independent RNA-seq datasets.
Foreground list consisted of 2000 genes with the smallest p-values.
Mean values are depicted as an “x”.
“*” indicates that 1 has been added to all values before log transformation.](
../analysis/fig2_ORAfdr.png "Figure 2A shows that correcting the FDR bug slightly reduces the number of significant
sets.")

![**Figure 2B. Impact of the FDR problem is worse for shorter foreground gene lists.**
Mean Jaccard index between original and corrected analysis were calculated for seven RNA-seq datasets with gene sets
from GO, Reactome, Cellmarkers and TFT GTRD; whiskers represent the standard deviation.](
../analysis/fig2_ORAsz.png "Figure 2B shows that the impact of the FDR bug is worse when
foreground gene lists are shorter than 500.")

@@

Simulated differential expression profiles with *a priori* changes to predetermined pathway
genes were analysed in parallel by clusterProfiler default and with mitigations for
problems described above.
In addition, these data underwent ORA with fora which does not suffer from these two
problems.
For comparison, fgsea was used to compare the performance of these two ORA methods against a
prototypical FCS method.
These methods were tested with different levels of introduced variation and the
precision, recall and F1 scores were recorded (**Figure 3A**).
Mitigating the background problem improved mean recall (by 2.9%), however it also caused
a decrease in mean precision (1.3% lower).
Addressing the FDR problem improved mean precision by 11.7%, but mean recall dropped by 16.6%.
Addressing both problems improved mean precision by 9.7% but decreased mean recall by 11.7%.
Under these conditions, the FDR problem affected results more severely than the background
problem.
Results obtained for clusterProfiler after implementing the two mitigations were identical to
fora to three significant figures.
fgsea recorded mean precision of 0.939, similar to the ORA methods, but recall was far superior,
with fgsea scoring 44.9% higher than default clusterProfiler and 64.2% higher than fora.
As a result, fgsea's overall accuracy as defined by F1 index was 35.4% higher than fora and
28.9% higher than default clusterProfiler.

When looking at the accuracy profile at different levels of added noise (**Figure 3B**),
ORA techniques show high precision when the added noise is low (<0.2), but
precision drops at higher levels of added noise.
The rate of drop in precision was more severe for approaches that suffer from the FDR problem.
The precision of fgsea was relatively stable despite high levels of added noise, ranging from
0.95 to 0.91.
The recall of fgsea was superior to all ORA approaches across all levels of added noise.

![**Figure 3A. Impact of correcting these two problems on accuracy of simulated expression
data.**
(A) Precision, recall and F1 scores across simulations.
CP; clusterProfiler, CP BG fix; clusterProfiler with a workaround to the background problem,
CP FDR fix; clusterProfiler with a workaround to the FDR problem,
CD BG and FDR fix; clusterProfiler with mitigations for both background and FDR problems,
fora; ORA function of the fgsea package,
fgsea; an FCS method similar to GSEA.
Mean values are depicted as an “x”.
](../analysis/sim/fig3_bars.png "Figure 3A shows that fora is more
precise than clusterProfiler, and fgsea has higher recall than ORA methods.")

@@

To understand whether FCS has better sensitivity than ORA in a real setting, we conducted a
systematic downsampling of a gene expression dataset consisting of normal and tumour samples
of 37 lung cancer patients.
In the full set of patient samples, fora (ORA) yielded 375 significantly differentially
regulated Reactomes, while fgsea (FCS) gave 480 (FDR<0.05).
Fgsea yielded consistently more significant pathways than fora in downsampled data
(**Figure 4A**).
Fgsea also identified a larger proportion of pathways in downsampled data compared to fora.
For example at a sample size of 5, fgsea identified 73.3% (median) of pathways from the full set as
compared to 41.5% for fora, confirming better recall for FCS over ORA.
It was noted that fgsea identified more pathways in downsampled data that were
inconsistent with the full dataset as compared to fora (**Figure 4B**).
However, inconsistent pathways were a smaller proportion of all findings for fgsea as
compared to fora (**Figure 4C**), indicating an overall lower false positive rate for FCS
compared to ORA methods.

![**Figure 4. Downsampling confirms higher sensitivity for FCS over ORA.**
(A) Violin plot showing the number of significant pathways identified in downsampled data that are consistent with
findings in the full dataset, using FCS (fgsea) and ORA (fora) methods.
Downsampling was repeated 50 times.
The blue and red horizontal lines indicate the number of significant pathways identified
in the full set of 37 patients using fgsea and fora respectively.
(B) The number of significant pathways identified in downsampled data that are not
consistent with the full dataset.
(C) The proportion of pathways in downsampled data that are inconsistent with the full
dataset.](
../analysis/fig4.png "Figure 4A shows that fgsea is more sensitive than fora. Figure 4B shows fgsea has higher
absule number of false positives, but Figure 4C shows that as a proportion of all findings fgsea has lower false
positive rate than fora.")

@@

The background problem came to our attention when a doctoral student in our department
was puzzled by results obtained from clusterProfiler when using a custom set of pathways.
This exact issue was the subject of several posts to online bioinformatics forums dating to
~2020 where confused users were wondering why the number of background genes changed so much.
The effect of excluding genes without functional annotations causes the background list to appear
smaller than it would otherwise be, causing enrichment scores to appear smaller.
The results shown here indicate the background problem affects smaller gene set libraries
like KEGG and HPO more severely as compared to Cellmarkers and GO libraries which describe
a larger proportion of all genes.
ClusterProfiler maintainers recently provided an option to prevent the loss of unannotated
genes from the background `options(enrichment_force_universe = TRUE)`, however this feature
isn't yet described in the official documentation.

@@

So end-users can explore the effect of these two problems on their work, we have developed
an R/Shiny web application (see "Availability of data and materials" below).
Users can upload their foreground and background lists, select their preferred gene set
library and compare results from default and corrected analyses.

ClusterProfiler is not the only tool to suffer from these problems, they are widespread
(**Table 2**). Still, there are some tools free of such problems.
For example, fora is a good option for R-based workflows, however the function does not calculate
the fold enrichment score,
so users will need to do it themselves [@Korotkevich2016-gd].
ShinyGO is an excellent alternative for web based analysis [@Ge2020-lu].
It doesn't suffer from the two issues described here, it is easy to use,
and has a high degree of reproducibility thanks to its stand-alone versioned docker
container option for local execution.
One problem that all of these tools listed in Table 2 have is that a background list is optional.
Sometimes this feature is hidden under "advanced options".
Given the strong biological and technical biases that are present in contemporary omics such as single cell
and spatial transcriptomics, it should be a mandatory step for users to provide a custom background list.

@@

These issues together with previously mentioned pitfalls such as lack of background
correction and methodological misreporting [@Timmons2015-ki;@Wijesooriya2022-fw], suggest
that if the data can be scored/ranked, methods like FCS are strongly recommended over ORA.
ORA is better suited to other cases where genes are classified into binary groups.
For example when investigating the over-represented categories in gene expression modules
[@Langfelder2008-hu], or with variant-harbouring genes associated with a disease [@White2019-ps].
In these cases, we would recommend using a tool from Table 2 that reports FDR values and enrichment scores,
and avoids background and FDR problems.
Doubts also arise when choosing options and parameters for ORA.
Most important is to use a background gene list that reflects the true universe of genes detected robustly in the
assay.
Secondly, significance of enrichment results must be based on the FDR-corrected p-values.
Selection criteria for foreground genes is another crucial choice; for this we endorse the approach of
Tarca et al, [@Tarca2013-xh] which accommodates cases where relatively few individual genes meet the FDR threshold.
Some ORA tools have a filter for gene set size.
We’d recommend a minimum size of 5-10 in order to avoid missing small gene sets that exhibit strong enrichments.
Interpret with caution enrichment results where the intersection of input gene list and pathway is ≤3 genes.
A maximum gene set filter is not required if end results are interpreted using enrichment scores together with FDR
values.
The choice of pathway database also influences the conclusions that can be drawn [@Mubeen2019-gw;@Karp2021-ue].
Our suggestion is to use comprehensive and open-source pathway databases such as Reactome and Gene Ontology
Biological Process.
Proprietary databases like KEGG and Ingenuity are less comprehensive, have restrictive conditions for non-academic
users and do not provide public access to historical database versions which inhibits future reproducibility.
